---
title: "What a coefficient tells you"
subtitle: "Why a SAOM parameter is not an effect size"
---

```{r setup01}
#| include: false
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
CRE <- "#2166ac"
MNT <- "#d73027"
```

## What we want from a model result

Interpreting a fitted model should let us say something fair and meaningful about
it — going beyond the direction and significance of a parameter, and relating the
result back to descriptives of the data that inform our understanding.

In an ordinary linear model this is nearly automatic. A coefficient is a change in
the outcome per unit of the predictor, in the outcome's own units, and it means the
same thing regardless of the values of all other observations.

A SAOM coefficient is a perfectly well-defined quantity too: a difference in
log-odds within a choice set, on the scale of the objective function. What changes
is that it no longer doubles as an answer to "how much", because four features
stand between it and the outcome scale:

1. **A non-linear link.** The same coefficient implies different changes in the
   outcome depending on the baseline.
2. **Interactions.** *Ceteris paribus* becomes hard to state given the complex
  hierarchies between effects, and harder again when predictors are collinear.
3. **Dependence between observations and over time.** Observations are not
   independent units.
4. **A latent decision sequence.** What the model describes — who considers changing
   which tie, and when — is never observed.

The fourth has no counterpart in cross-sectional network models, and it is the
reason there will turn out to be more than one average marginal effect rather than
a single one. Taken together these aspects make it harder to link back the results
to any descriptives of the data you have made. It's also makes it difficult to
interpret their actual impact on decisions and the observed data.

It is worth being clear that none of these complications are new:
they are already in the model, and going beyond constant log-odds coefficients is
what brings them to light. Effect-size questions are what make them unavoidable —
to state an effect size you have to say what is being conditioned on, what is
being averaged over, and on which scale. The model has always worked this way;
asking it for effect sizes is what makes that structure visible.

## What a coefficient does answer

That list is what the coefficient scale will not do on its own. It is worth being
equally concrete about what it will. Below is one school class — `12b`, the first
of the three used in Chapter 4 — fitted with reciprocity, transitivity, a
primary-school indicator and gender.

```{r coef-table}
#| echo: false
co  <- readRDS("results/estimation_results.rds")$coefs
m   <- co[co$model == "model_2" & co$class_id == "12b", ]
lab <- c(friendship_rate1               = "rate (period 1)",
         friendship_density_eval        = "outdegree (density)",
         friendship_recip_eval          = "reciprocity",
         friendship_transTrip1_eval     = "transitive triplets",
         friendship_X_primary_eval      = "same primary school",
         friendship_altX_sex_m_eval     = "male alter",
         friendship_egoX_sex_m_eval     = "male ego",
         friendship_egoXaltX_sex_m_eval = "both male",
         friendship_unspInt_eval        = "reciprocity × transitivity")
data.frame(effect   = unname(lab[m$effect]),
           estimate = round(m$est, 3),
           SE       = round(m$se, 3))
```

Each estimate is a difference in satisfaction between two networks that differ by
one unit of the corresponding statistic. Comparing two mechanisms is therefore a
subtraction, and questions that sound quantitative can be put directly.

Take the situation in @fig-two-alternatives. A boy $i$ has been nominated by $j$, a
girl, and is not tied to $k$, another boy. Two of the alternatives open to $i$ at
his next opportunity are to reciprocate $j$'s nomination, or to nominate $k$. Which
does the model say he prefers?

![A stylized decision situation. (a) the starting network: $j$ has nominated $i$,
and $k$ is unconnected; node colour is gender, so $i$ and $k$ are boys and $j$ is a
girl. (b) $i$ reciprocates, giving a mutual tie to a girl. (c) $i$ nominates $k$,
giving an unreciprocated tie to a
boy.](figs/two_alternatives.svg){#fig-two-alternatives width="70%"}

Both are tie creations, and in a network this bare neither closes a two-path, so
the outdegree term, ego's own gender and the structural terms enter both
alternatives alike. Take $j$ and $k$ to be alike on primary school too, and what is
left is reciprocity on one side against the two alter-gender terms on the other:

| alternative | terms that differ | total |
|---|---|---|
| (b) reciprocate $j$ | reciprocity $+2.05$ | $+2.05$ |
| (c) nominate $k$ | male alter $-0.70$, both male $+1.73$ | $+1.03$ |

The model puts reciprocating about one unit ahead. Nothing in that reasoning is
illegitimate: on the satisfaction scale, coefficients compare.

What the table cannot tell you is whether a gap of about one unit is
distinguishable from zero. It gives a standard error for each coefficient, but the
quantity compared here combines three of them, and the standard error of a
combination depends on how those three estimators covary — which a column of
standard errors does not contain. That is what `test_parameter()` is for: it tests
a linear combination of parameters using the full covariance matrix. Here it would
matter, since reciprocity is estimated loosely in a class this size.

So coefficients do answer comparative questions about satisfaction, and answer them
properly. What they will not tell you is how much more likely $i$ actually is to
reciprocate — a question on a different scale, and the one the rest of this material
is about.

## Already visible in the simplest case

Take a logistic regression with one predictor, a coefficient of 1, and an intercept
$C$. Only a null model would be simpler in this family, and already the coefficient does not translate into a single probability change.

```{r}
#| label: fig-logistic
#| fig-cap: "The same coefficient, two different probability changes. Left: a one-unit increase in $x$ from two starting points. Right: the same one-unit increase under two values of the intercept $C$. After @Glasgow2022, Fig. 1."
#| fig-width: 8.5
#| fig-height: 3.8
#| echo: false
xs <- seq(-4, 4, length.out = 400)

pa <- data.frame(x = xs, p = plogis(xs), panel = "(a) same step, different start")
seg_a <- data.frame(x0 = c(-1, 1), x1 = c(0, 2),
                    panel = "(a) same step, different start")
seg_a$p0 <- plogis(seg_a$x0); seg_a$p1 <- plogis(seg_a$x1)

pb <- rbind(
  data.frame(x = xs, p = plogis(xs),       panel = "(b) same step, different C", C = "C = 0"),
  data.frame(x = xs, p = plogis(xs + 1.5), panel = "(b) same step, different C", C = "C = 1.5"))
seg_b <- data.frame(x0 = c(0, 0), x1 = c(1, 1), shift = c(0, 1.5),
                    panel = "(b) same step, different C")
seg_b$p0 <- plogis(seg_b$x0 + seg_b$shift)
seg_b$p1 <- plogis(seg_b$x1 + seg_b$shift)

seg <- rbind(seg_a[, c("x0", "x1", "p0", "p1", "panel")],
             seg_b[, c("x0", "x1", "p0", "p1", "panel")])
seg$lab <- sprintf("%.2f", seg$p1 - seg$p0)

ggplot() +
  geom_line(data = pa, aes(x, p), colour = "grey25") +
  geom_line(data = pb, aes(x, p, group = C, colour = C)) +
  geom_segment(data = seg, aes(x = x0, xend = x1, y = p0, yend = p0),
               colour = "grey55", linetype = "dotted") +
  geom_segment(data = seg, aes(x = x1, xend = x1, y = p0, yend = p1),
               colour = CRE, linewidth = 0.8) +
  geom_text(data = seg, aes(x = x1 + 0.22, y = (p0 + p1) / 2, label = lab),
            hjust = 0, size = 3.2, colour = CRE) +
  scale_colour_manual(values = c("C = 0" = "grey25", "C = 1.5" = MNT), name = NULL) +
  facet_wrap(~ panel) +
  coord_cartesian(xlim = c(-4, 4.6), ylim = c(0, 1)) +
  labs(x = "x", y = "probability") +
  theme(legend.position = "bottom")
```

On the left, a one-unit increase in $x$ raises the probability by about 0.23 near
the middle of the curve and about 0.15 higher up. Same coefficient, different
effect. On the right, the same one-unit increase evaluated under two intercepts
again gives two different answers.

Odds ratios sit in a similar position. An odds ratio *is* constant — that is its
appeal — but constant on the odds scale and multiplicative, so it leaves open how
far the probability moves. Which matters when the claim being made is about
probabilities or how it affects the observed outcome.

## The same question about comparisons

Comparing coefficients raises the issue again. Three comparisons come up often:

| comparison | question | example |
|---|---|---|
| one data set, one model | which mechanism matters more? | reciprocity vs closure |
| one data set, several models | does the effect change when we add a control? | mediation, moderation |
| several data sets, one model | how much does the effect vary across settings? | meta-analysis |

The first row runs into something elementary before any of that. What enters the
objective function is the product $\beta_k s_{ki}$, so how large a coefficient comes
out is tied to the range its statistic takes. Reciprocity is an indicator moving
between 0 and 1; a count of transitive triplets can move by ten. Reading the two
coefficients against each other as a statement about which mechanism matters more
compares steps of one on scales whose steps are not the same size. This is the
difficulty that standardised coefficients handle in linear regression, and for which
non-linear models have no clean equivalent — though partial remedies exist, and
@AMAND discusses semi-standardised effects for the SAOM.

All three rows then run into a second problem. Each is relatively straightforward on
the coefficient scale only if the
scale itself is the same on both sides of the comparison, and in a non-linear model
it need not be: a coefficient is identified relative to how much variation the model
leaves unexplained. Two familiar habits also do not carry over — there is no 
constant residual variance, and no useful notion of explained variance.

The mechanism is worth stating precisely, because its direction is easy to lose.
Leaving a predictor out *increases* the unexplained variation. The scale is
implicitly fixed, so the remaining coefficients **contract** to absorb it — meaning
coefficients are systematically smaller in models with fewer predictors, and grow
as predictors are added. This is the same rescaling long discussed for binary
logistic regression [@Mood2010].

The SAOM has a second absorber with no binary-logit counterpart: the rate parameter
can inflate, so simulated actors take more ministeps to cover the same observed
change. In one simulation experiment, dropping an orthogonal covariate attenuated
the structural coefficients by roughly 23% while inflating the rate parameter by
roughly 43%.

The third row has an established response — meta-analysis, with a long tradition of
standardised cross-study effect sizes and, for SAOMs, a worked-out random-effects
treatment: see section 12.3 of the RSiena manual [@SienaManual], and the two-stage
population model in @AMAND. Chapter 4 puts it to use. Two things from above bear on
it, and they pull in opposite directions.

One is an artefact. Specifications across studies are rarely identical, and when the
results being synthesised are other people's you take the specifications you are
given; since what a model leaves unexplained is folded into its coefficients, the
numbers entering a pool need not sit on a common scale.

The other is not really a problem, though it is easy to mistake for one. Settings
and actors genuinely differ in how predictable behaviour is, and where they do, the same
underlying tendency really does translate into different outcome probabilities. 
An effect size that comes out differently there is reporting something true, 
not failing to control for something. Chapter 6 says what a comparison of this 
kind can and cannot claim.

## Where this is going

What the rest of the workshop material offers is a second quantity to report alongside the
coefficient, on the probability scale: how much does a mechanism change the probability
that a tie is created, or maintained? The coefficient stays what it is and keeps its own
uses; this is an additional estimand, aimed at readers who want to reason on the
probability scale, or to set model results next to the descriptive comparisons one
can compute from two waves directly — with the limits of those comparisons stated
as they arise.

Unlike a coefficient, such a quantity has to be *defined* rather than read off:
what is held fixed, what is varied, and over which situations it is averaged.
That is the work of Chapter 2.

One expectation to set now. Moving to the probability scale gives a common unit
and a directly interpretable quantity. It does not by itself give comparability
across scales, and Chapter 6 will expand on this.
