---
title: "Interpreting a SAOM"
subtitle: "From coefficients to probabilities, one decision at a time"
format:
  html:
    toc: true
    toc-depth: 2
    df-print: kable
execute:
  warning: false
  message: false
---

## Setup — run once

Everything here needs the `sienaMargins` branch of RSiena, which is where
`predict()` and `marginalEffects()` live, plus `ggplot2` for the figures.

**These chunks do not run when the document is rendered.** Run whichever one
matches your setup by hand, then restart R.

```{r install}
#| eval: false
install.packages(c("remotes", "ggplot2"))
remotes::install_github("stocnet/rsiena@sienaMargins")
```

If you keep this project's dependencies in `renv` instead:

```{r install-renv}
#| eval: false
install.packages("renv")
renv::install(c("stocnet/rsiena@sienaMargins", "ggplot2"))
```

Check that it worked — `predict()` needs to accept a targets object:

```{r check}
#| eval: false
packageVersion("RSiena")
library("RSiena")
exists("make_predict_targets")   # TRUE on the sienaMargins branch
```

Afterwards we can setup 

```{r setup}
#| include: false
library(RSiena)
library(ggplot2)
set.seed(42)
theme_set(theme_minimal(base_size = 12))
CRE <- "#2166ac"   # creation
MNT <- "#d73027"   # maintenance
```

## The model we are trying to interpret

Take the `s50` data that ships with RSiena: 50 pupils, three waves of friendship
nominations, plus smoking and drinking. Two actor covariates, so the model has
both kinds of effect — structural effects belonging to a single alternative, and
covariate effects that move an actor's whole choice set at once.

```{r fit}
mynet   <- as_dependent_rsiena(array(c(s501, s502, s503), dim = c(50, 50, 3)))
smoke   <- as_covariate_rsiena(s50s[, 1])
alcohol <- as_covariate_rsiena(s50a[, 1])
mydata  <- make_data_rsiena(mynet, smoke, alcohol)

myeff <- make_specification(mydata)
myeff <- set_effect(myeff, transTrip1, verbose = FALSE)
myeff <- set_effect(myeff, egoX, covar1 = "smoke",   verbose = FALSE)
myeff <- set_effect(myeff, egoX, covar1 = "alcohol", verbose = FALSE)
myeff <- set_effect(myeff, altX, covar1 = "alcohol", verbose = FALSE)

## A user-specified interaction of reciprocity and transitivity.  Its change
## statistic is the product of the two components' -- which matters in Step 8.
myeff <- set_interaction(myeff, c(recip, transTrip1), verbose = FALSE)

myalgo <- set_algorithm_saom(cond = FALSE, n3 = 1000, seed = 42)
ans <- siena(data = mydata, effects = myeff, control_algo = myalgo,
             batch = TRUE, verbose = FALSE)
```


```{r coefs}
data.frame(effect   = names(coef(ans)),
           estimate = round(coef(ans), 3),
           SE       = round(sqrt(diag(vcov(ans))), 3),
           row.names = NULL)
```

Reciprocity is large and positive. **But how large?** The coefficient is a
change in a log-odds inside a choice among 50 alternatives. It is not a
probability, not a rate, and not easily comparable to the reciprocity coefficient of
another model on another network.

This document walks from that coefficient to quantities that are (more so).

## Step 1 — the descriptive difference

Three simple statistics we can count directly between two observed waves, with no model.

```{r descriptives}
as_tie <- function(m) { m[m == 10] <- NA; diag(m) <- NA; m }
x1 <- as_tie(s501); x2 <- as_tie(s502)
ok <- !is.na(x1) & !is.na(x2)

p_rec <- mean(x2[ok & t(x1) == 1] == 1)
p_not <- mean(x2[ok & t(x1) == 0] == 1)

data.frame(
  quantity = c("tie creation (non-ties at t1 that appear)",
               "tie maintenance (ties at t1 that survive)",
               "P(tie at t2 | reciprocated at t1)",
               "P(tie at t2 | not reciprocated at t1)",
               "descriptive difference in probability"),
  value = round(c(mean(x2[ok & x1 == 0] == 1),
                  mean(x2[ok & x1 == 1] == 1),
                  p_rec, p_not, p_rec - p_not), 3))
```

The last row is the **descriptive difference in probability**: are ties more prevalent where
the condition holds? Any researcher can compute it directly from two waves.

It is also a gap that reflects **all mechanisms at once**. Dyads that reciprocate
are also more transitively embedded, and involve pupils who differ in other ways.
Nothing here separates a reciprocity mechanism from any of that.

Note too how different tie creation and tie maintenance are. Hold on to that.

## Step 2 — what the model sees

Before asking the model anything, it is worth looking at what it works from. The
**change statistics** describe every alternative open to every actor: one row per
ego–alter pair.

```{r change-stats}
cs <- getChangeStatistics(ans, mydata, effects = myeff, depvar = "mynet")

head(cs[cs$ego != cs$choice, ], 4)
```

Each row says: is this alternative reciprocated, how many two-paths does it have,
do the two pupils drink alike - equivalently how much would the effect statistic
differ if the tie to that alternative would exist vs if it would not exist. 
The `mynet_recip` column at period 1 is exactly the condition from Step 1 — 
so the descriptive difference can be recomputed from it, without the matrix algebra:

```{r descr-from-cs}
p1 <- cs[cs$period == 1 & cs$ego != cs$choice, ]
tie2 <- as_tie(s502)[cbind(p1$ego, p1$choice)]      # was there a tie at wave 2?

c(reciprocated     = mean(tie2[p1$mynet_recip == 1] == 1, na.rm = TRUE),
  not_reciprocated = mean(tie2[p1$mynet_recip == 0] == 1, na.rm = TRUE))
```

The same two numbers as Step 1. That is the point: the condition you can count by
hand and the condition the model conditions on are the same column. Everything
from here on splits these rows in different ways.

::: {.callout-note collapse="true"}
## Change statistics and toggle statistics differ in sign
The toggle statistic of Chapter 2 is what `flip = FALSE` returns, under the name
*change contribution*:

```{r contributions}
cc <- getChangeStatistics(ans, mydata, effects = myeff, depvar = "mynet",
                          flip = FALSE)

## an existing tie that is also reciprocated, so the difference is visible
i    <- which(cs$mynet_density == -1 & cs$mynet_recip == 1)[1]
cols <- c("mynet_density", "mynet_recip", "mynet_transTrip1")

rbind(`change statistic`    = cs[i, cols],
      `toggle statistic` = setNames(cc[i, paste0(cols, "_eval")], cols))
```

Remember the distinction: the **toggle statistic** is what toggling would *do*, so
dropping a reciprocated tie loses the reciprocity and it comes out negative. The
**change statistic** is the same information as a property of the alternative —
this alter *does* nominate ego, so it is positive. Conditioning uses the change
statistic, which is why "condition on `recip`" means "is the reverse tie there",
not "what would toggling do".

Density is the exception: it is left on the toggle scale in both, so that its
`+1` / `-1` marks the direction of the decision. As a change statistic it would be
`+1` throughout, since a tie adds one to density whichever way the decision runs —
which would be true but useless for telling creation from maintenance.

These describe ego–alter pairs, which is what we want here. They can also be
added up into ordinary network descriptives — the rows with `density == -1` are
the ties that exist, so:

```{r recip-index}
ties <- p1[p1$mynet_density == -1, ]

c(ties              = nrow(ties),
  reciprocated      = sum(ties$mynet_recip),
  reciprocity_index = round(mean(ties$mynet_recip), 3))
```

which is the wave-1 network's reciprocity index, computed without touching the
adjacency matrix. This works for effects that are linear in the focal tie. It
does **not** work for transitivity, where one tie takes part in several triplets
and so would be counted more than once.
:::

## Aside — the same statistics along simulated ministeps

The observed waves are only the starting points. `dynamic = TRUE` returns the
change statistics along the model's own simulated ministep sequences, with
`chain` and `ministep` columns added:

```{r dynamic-cs}
dyn <- getChangeStatistics(ans, mydata, effects = myeff, depvar = "mynet",
                           dynamic = TRUE, algorithm = myalgo, n3 = 4)

dim(dyn)
head(dyn[, c("chain", "ministep", "ego", "choice", "mynet_recip")], 3)
```

Four chains already give over a hundred thousand rows. Do the decision
situations drift much as a chain runs?

```{r drift-plot}
#| fig-width: 8
#| fig-height: 3.2
d1  <- dyn[dyn$period == 1, ]
agg <- aggregate(cbind(mynet_recip, mynet_transTrip1) ~ ministep, d1, FUN = mean)

long <- rbind(
  data.frame(ministep = agg$ministep, statistic = "reciprocated",
             value = agg$mynet_recip),
  data.frame(ministep = agg$ministep, statistic = "two-paths",
             value = agg$mynet_transTrip1))

ggplot(long, aes(ministep, value)) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_smooth(se = FALSE, colour = CRE, linewidth = 0.8) +
  facet_wrap(~ statistic, scales = "free_y") +
  labs(x = "ministep within the period",
       y = "mean over the alternatives on offer")
```

Fairly flat: the mix of alternatives actors face does not change dramatically
between the start and the end of a period here. That is reassuring for everything
that follows, where the observed contexts stand in for the whole path — but it is a
property of this network, not a guarantee, and worth checking when the network
moves a lot.

You can also keep the contributions from the fit itself, with
`set_output_saom(returnChangeContributions = TRUE)` in `siena()`. That stores
every ministep of every phase-3 simulation, so it grows with actors, effects and
simulations and becomes expensive in memory for larger models. `predict()` and
`marginalEffects()` reach the same quantities internally but re-simulate in
batches, so you do not normally need to hold them all at once.

## Step 3 — the same comparison inside the model

The descriptive difference compares two *observed networks*. The SAOM does not
model that jump directly. It models the process **between** the waves as a
sequence of **ministeps**: one actor gets an opportunity, looks at every
alternative open to it, and toggles one tie — creating a new one, or keeping or
dropping one it has.

So every quantity below sits at one ministep and asks: for this actor, facing
this alternative, what happens next?

`predict()` gives those probabilities. Because it works from the change
statistics, we can split them by features of the decision situation.

```{r predict}
ptg <- make_predict_targets(ans, effects = myeff, depvar = "mynet",
                            type = "tieProb", level = "period",
                            includeDefaults = FALSE)
ptg <- set_condition(ptg, recip,   name = "by_recip",   verbose = FALSE)
ptg <- set_condition(ptg, density, name = "by_context", verbose = FALSE)
ptg <- set_condition(ptg, c(density, recip), name = "by_both", verbose = FALSE)

pred <- predict(ans, mydata, targets = ptg,
                control_uncertainty = set_postest_uncertainty_saom(mode = "delta"))
```

These are tie probabilities — the probability the tie **exists** once the actor
has taken its ministep — which is what `type = "tieProb"` asks for. Split by
whether the reverse tie is already there:

```{r by-recip}
pred$by_recip
```

This is the Step 1 comparison again, now from the model and at the ministep
rather than between two waves. It reflects **all mechanisms at once** in exactly
the same way: nothing here holds the decision situation fixed.

::: {.callout-note collapse="true"}
## Which actors are being compared?
Two separate things affect who enters these averages, and they are easy to run
together.

The first is already taken care of. Averaging is normalised per actor by default
(`egoNormalize = TRUE`): each actor's alternatives are averaged first, and those
per-actor values are then averaged. So a pupil with forty open alternatives does
not count for more than one with five.

The second is not. Conditioning on `recip` splits the rows into two strata, and
each is averaged over the actors who *have* that kind of alternative — which is not
the same set of pupils. Some have no reciprocated alternative at all, so they enter
the `recip = 0` average only, and the gap subtracts two means taken over different
people.

Differencing inside each actor's own choice set first, and averaging only over
actors with both kinds, avoids that. `level = "ego"` returns one value per actor
per stratum, which is what makes it possible:

```{r ego-level}
etg <- make_predict_targets(ans, effects = myeff, depvar = "mynet",
                            type = "tieProb", level = "ego",
                            includeDefaults = FALSE)
etg <- set_condition(etg, recip, name = "by_recip", verbose = FALSE)
eg  <- as.data.frame(predict(ans, mydata, targets = etg,
         control_uncertainty = set_postest_uncertainty_saom(enabled = FALSE)))

pooled <- as.data.frame(pred$by_recip)

do.call(rbind, lapply(sort(unique(eg$period)), function(pd) {
  e   <- eg[eg$period == pd, ]
  p   <- pooled[pooled$period == pd, ]

  ## pupils who face both kinds of alternative in this period
  tab  <- table(e$ego, e$mynet_recip)
  both <- rownames(tab)[rowSums(tab > 0) == 2]
  e    <- e[e$ego %in% both, ]

  data.frame(period     = pd,
             pupils     = nrow(tab),
             with_both  = length(both),
             pooled     = p$tieProb[p$mynet_recip == 1] -
                          p$tieProb[p$mynet_recip == 0],
             within_ego = mean(sapply(split(e, e$ego), function(d)
                            d$tieProb[d$mynet_recip == 1] -
                            d$tieProb[d$mynet_recip == 0])))
}))
```

`with_both` is how many pupils the balanced version can use, and it changes between
periods. The two figures agree to about a thousandth here, so nothing in Step 3
turns on the choice — but that is a fact about this network, not a general one. It
matters more when actors differ a lot in who is available to them.

This only works for statistics that vary across an actor's alternatives.
Ego's alcohol is the same for all of them, so that question can only be asked
across actors.
:::

## Step 4 — creation and maintenance are different decisions

The creation and maintenance margins are two different questions, and on the
tie-probability scale they have very different answers: an existing tie is very
likely to still be there after one ministep, an absent one very unlikely to have
appeared. The `density` column tells the two apart — `+1` where the actor is
deciding whether to create a tie, `-1` where the decision is whether to maintain
one.

```{r by-context}
pred$by_context
```

Reciprocity is strongly associated with the tie already existing. So how much of
the Step 3 gap is really "these dyads were already tied"?

```{r by-both}
## the reciprocity gap: P(tie | reciprocated) - P(tie | not reciprocated)
gap <- function(d) d$tieProb[d$mynet_recip == 1] - d$tieProb[d$mynet_recip == 0]

pooled <- as.data.frame(pred$by_recip)
b <- as.data.frame(pred$by_both)
b$decision <- ifelse(b$mynet_density > 0, "tie creation", "tie maintenance")

## the same gap taken separately within each kind of decision
gaps <- do.call(rbind, lapply(split(b, list(b$period, b$decision), drop = TRUE),
  function(s) data.frame(period = s$period[1], decision = s$decision[1],
                         reciprocity_gap = gap(s))))

out <- rbind(do.call(rbind, lapply(split(pooled, pooled$period), function(s)
               data.frame(period = s$period[1], decision = "unconditional",
                          reciprocity_gap = gap(s)))),
             gaps)
out$reciprocity_gap <- round(out$reciprocity_gap, 3)
out <- out[order(out$period), ]
rownames(out) <- NULL
out
```

Read the first row against the other two. Taking both kinds of decision together
gives a **larger** gap than either kind on its own, so it is clearly not an
average of them.

The reason is that reciprocated and unreciprocated alternatives are not the same
kind of decision. Where the reverse tie already exists, the tie itself usually
exists too, so those are mostly maintenance decisions with a high probability.
Where it does not, the tie usually does not exist either, so those are mostly
creation decisions with a low probability. Comparing the two therefore compares
existing ties against absent ones, with reciprocity varying alongside.

Most of the apparent reciprocity gap in Step 3 was the state of the tie itself.

## Step 5 — the adjusted effect

Splitting by density accounts for one thing the comparison was picking up. The
model contains others: transitivity, drinking, smoking, and the reciprocity ×
transitivity interaction. A dyad that reciprocates is usually also more
transitively embedded.

To isolate reciprocity we hold the decision situation fixed: same actor, same
alternative, every other statistic where it actually is, and only reciprocity
switched from 0 to 1. That contrast is the **first difference**.

One decision is not an answer, though, so the first differences are averaged
over all the decision situations in the data. That average is the **average
marginal effect**, usually shortened to **AME** — the standard name for "the
effect of switching this on, averaged over everyone it could apply to". It is
what the table below reports.

```{r first-diff}
mtg <- make_marginal_targets(ans, effects = myeff, depvar = "mynet",
                             type = "tieProb", level = "period",
                             condition = "density", includeDefaults = FALSE)
mtg <- set_target(mtg, recip, contrast = c(0, 1), name = "recip_fd",
                  verbose = FALSE)
## Perturbing recip also moves the recip x transTrip1 interaction, because that
## interaction's change statistic is the product of its components'.  Declared
## once, applied wherever it is reached.
mtg <- set_dependency(mtg, unspInt ~ recip:transTrip1, verbose = FALSE)

fd <- marginalEffects(ans, mydata, targets = mtg,
        control_uncertainty = set_postest_uncertainty_saom(mode = "delta"))
fd
```


## Step 6 — how much the adjustment moves it

Step 4 left three quantities in play: the unconditional gap, the gap within each
kind of decision, and — from Step 5 — the adjusted first difference. The first was
already shown to be an artefact of pooling two different decisions, so what is
worth comparing is the last two, within each margin:

```{r compare}
f <- as.data.frame(fd)
f$decision <- ifelse(f$mynet_density > 0, "tie creation", "tie maintenance")

cmp <- merge(gaps, f[, c("period", "decision", "firstDiff")],
             by = c("period", "decision"))

data.frame(period     = cmp$period,
           decision   = cmp$decision,
           unadjusted = round(cmp$reciprocity_gap, 3),
           adjusted   = round(cmp$firstDiff, 3))
```

Same model, same ministep, same probability scale. The two differ in one
respect only: whether the decision situation is held fixed. So the difference
between them is what the raw comparison was picking up from other mechanisms.

It does not always point the same way. For tie creation the unadjusted
comparison overstates reciprocity, as expected. For tie maintenance it
*understates* it:
the adjusted effect is larger. That is not a mistake. Mutual ties differ from
asymmetric ties in more than reciprocity — they are also more transitively
embedded, and the reciprocity × transitivity interaction here is negative, which
holds their raw survival advantage down. Holding the decision situation fixed
puts it back.

Adjustment does not only ever shrink things.

::: {.callout-important}
## What not to compare directly
The Step 1 descriptive difference is **period-level**: it compares two waves.
The AME is **per-opportunity**: it refers to one ministep. They are not on the
same footing, so the difference between them is not interpretable.

Summing the contrast along the ministeps the model implies between one wave and
the next does give a period-level quantity — the **accumulated effect** — but it
is a cumulative count rather than a probability, so it does not simply close the
gap either. Chapter 6 sets out what it does and does not support. For now, treat
Step 1 as an anchor for intuition rather than as a number to subtract from.
:::

Everything above is evaluated at the **observed** waves. The model can instead
average over its own **model-implied decision sequences** (`dynamic = TRUE`),
which uses the whole simulated path between waves rather than only the observed
states. Same functions, one extra argument.

## Step 7 — effects are not constant

An AME answers "how much, on average", and hides on whom and when. The
underlying contrasts are **context-conditional** — they depend on the decision
situation, and the average is only a summary of them. Stratifying on a feature
of that situation gives a **conditional AME**: the same average, computed
separately within each group of decisions, which shows the effect as a function
of that feature.

Here: what does one more two-path do, given how many are already there?

```{r shape}
stg <- make_marginal_targets(ans, effects = myeff, depvar = "mynet",
                             type = "tieProb", level = "period",
                             condition = c("density", "transTrip1"),
                             includeDefaults = FALSE)
stg <- set_target(stg, transTrip1, diff = 1, name = "tt", verbose = FALSE)
stg <- set_dependency(stg, unspInt ~ recip:transTrip1, verbose = FALSE)

shape <- as.data.frame(marginalEffects(ans, mydata, targets = stg,
          control_uncertainty = set_postest_uncertainty_saom(mode = "delta")))
shape <- shape[shape$mynet_density != 0, ]
shape$context <- ifelse(shape$mynet_density > 0, "creation", "maintenance")
```

```{r shape-plot}
#| fig-width: 9
#| fig-height: 5
ggplot(shape, aes(mynet_transTrip1, firstDiff, ymin = q_025, ymax = q_975,
                  colour = context, fill = context)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.8) +
  facet_grid(period ~ context, scales = "free_y", labeller = label_both) +
  scale_colour_manual(values = c(creation = CRE, maintenance = MNT),
                      guide = "none") +
  scale_fill_manual(values = c(creation = CRE, maintenance = MNT),
                    guide = "none") +
  labs(x = "two-paths already present", y = "effect of one more two-path",
       caption = "95% delta-method intervals")
```

Read it: is the line flat (a constant effect), rising (self-reinforcing), or
falling (saturating)? And does it do the same in both panels? A single average
tells you none of this — this is the nonlinear, context-dependent heterogeneity
that the average is a summary of.

The same idea with a binary feature: does reciprocity change what a two-path
does?

```{r mod}
#| fig-width: 8
#| fig-height: 4.5
mtg2 <- make_marginal_targets(ans, effects = myeff, depvar = "mynet",
                              type = "tieProb", level = "period",
                              condition = c("density", "recip"),
                              includeDefaults = FALSE)
mtg2 <- set_target(mtg2, transTrip1, diff = 1, name = "tt", verbose = FALSE)
mtg2 <- set_dependency(mtg2, unspInt ~ recip:transTrip1, verbose = FALSE)

mod <- as.data.frame(marginalEffects(ans, mydata, targets = mtg2,
        control_uncertainty = set_postest_uncertainty_saom(mode = "delta")))
mod <- mod[mod$mynet_density != 0, ]
mod$context <- ifelse(mod$mynet_density > 0, "creation", "maintenance")
mod$recip   <- factor(mod$mynet_recip, levels = c(0, 1),
                      labels = c("not reciprocated", "reciprocated"))

ggplot(mod, aes(recip, firstDiff, ymin = q_025, ymax = q_975,
                colour = context)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange() +
  facet_grid(period ~ context, scales = "free_y", labeller = label_both) +
  scale_colour_manual(values = c(creation = CRE, maintenance = MNT),
                      guide = "none") +
  labs(x = NULL, y = "effect of one more two-path")
```

## Step 8 — the interaction, as a second difference

The plot above suggests the two-path effect differs by reciprocity. The
**second difference** states that as one number with an interval: how much does
the effect of one more two-path change when the tie goes from unreciprocated to
reciprocated?

```{r second-diff}
sdt <- make_marginal_targets(ans, effects = myeff, depvar = "mynet",
                             type = "tieProb", level = "period",
                             condition = "density", includeDefaults = FALSE)
sdt <- set_second_diff(sdt,
                       list(transTrip1 = list(diff = 1),
                            recip      = list(contrast = c(0, 1))),
                       name = "tt_x_recip", verbose = FALSE)
sdt <- set_dependency(sdt, unspInt ~ recip:transTrip1, verbose = FALSE)

marginalEffects(ans, mydata, targets = sdt,
    control_uncertainty = set_postest_uncertainty_saom(mode = "delta"))
```

The dependency is declared once and reaches both steps of the second
difference, so the cross term falls out of the arithmetic rather than being
rebuilt by hand. And, as with everything else here, the answer differs between
tie creation and tie maintenance.

## Beyond this walkthrough

The walkthrough kept to one kind of target — a statistic that varies across ego's
alternatives — evaluated at the observed waves. Two things beyond that are worth
knowing about, neither of which was covered in the session.

**An ego statistic needs a different contrast.** Every target so far has been a statistic that varies across ego's alternatives —
reciprocity, two-paths. An effect that only depends on ego's covariate value does not: 
`egoX` for alcohol is the same number for every alternative this actor has, 
so it cannot be moved for one alter on its own. The perturbation has to shift 
ego's whole choice set at once, raising every
creation alternative and lowering every dissolution one.

That is worked out from the effect, so nothing extra has to be asked for. Setting a
target on an ego effect switches the perturbation to the ego-wide form and brings the
probability-mass contrasts along with it. Alcohol is a graded 1–5 scale rather than
an indicator, so the contrast to ask for is `diff = 1` — one unit more drinking —
rather than a $0 \to 1$ switch:

```{r ego-target}
etg <- make_marginal_targets(ans, effects = myeff, depvar = "mynet",
                             type = "tieProb", level = "period",
                             includeDefaults = FALSE)
etg <- set_target(etg, egoX, covar1 = "alcohol", diff = 1,
                  name = "alcohol_ego", verbose = FALSE)

marginalEffects(ans, mydata, targets = etg,
    control_uncertainty = set_postest_uncertainty_saom(mode = "bootstrap",
                                                       nsim = 200))
```

Three numbers instead of one, answering different questions. The first difference
(`firstDiff`) is the per-tie one: how much one focal friendship's probability
moves. The two probability-mass contrasts (`massCreation`, `massDissolution`) are
the totals across the whole choice set — how much probability moves into forming
*some* friendship and out of dropping *one*. Their difference 
is the effect on expected degree change, which for a sender effect is arguably a
better representation of its impact than the average first difference. Each
estimate carries its own interval, named after the quantity it belongs to.

One thing to keep in mind about `diff`: it shifts every actor by one unit from
wherever they start, so a pupil already at 5 is moved to 6 — a value the scale
never takes. Holding the contrast the same size for everyone is what makes the
average interpretable, but it does mean part of that average is evaluated off the
observed range. Worth remembering whenever a covariate has a hard ceiling or floor.

The perturbation mode can be overridden — `set_target(..., perturbType = "alter")`,
and `massContrasts` to force the extra columns on or off — though there is usually
no reason to, beyond something specific such as comparing ego and alter effects
directly.

**Averaging over model-implied ministeps.** `dynamic = TRUE` on the targets replaces
the observed-wave decision situations with the ones the fitted model says occur
between waves. It needs an algorithm to simulate with, which goes in `control_algo`:

```{r dynamic-me}
dtg <- make_marginal_targets(ans, effects = myeff, depvar = "mynet",
                             type = "tieProb", level = "period",
                             dynamic = TRUE, includeDefaults = FALSE)
dtg <- set_target(dtg, recip, contrast = c(0, 1), name = "recip_fd",
                  verbose = FALSE)
dtg <- set_dependency(dtg, unspInt ~ recip:transTrip1, verbose = FALSE)

marginalEffects(ans, mydata, targets = dtg,
    control_algo = set_postest_algo_saom(algorithm = myalgo, n3 = 100,
                                         verbose = FALSE))
```

Compare with the observed-wave answer from Step 5. When the two agree, the
observed-wave summary is not being driven by a few atypical situations.

The standard error here is still analytic, with one addition: averaging over
simulated chains means the chains themselves depend on the parameters, so a
correction for that is included alongside the usual delta-method term. It happens
by itself for a dynamic target.

## Where this leaves us

| question | quantity |
|---|---|
| are ties more prevalent where the condition holds? | descriptive difference (Step 1) |
| what does the model work from? | change statistics (Step 2) |
| what does the model say about one decision? | predicted probabilities (Step 3) |
| is that just creation versus maintenance? | split by margin (Step 4) |
| what does *this* mechanism contribute? | first difference, averaged into an AME (Step 5) |
| how much of the raw gap is the mechanism? | compare the two (Step 6) |
| is the effect constant? | conditional AMEs (Step 7) |
| do two mechanisms interact? | second difference (Step 8) |
