---
title: "A meta-analysis at the smallest useful size"
subtitle: "Three classes, one specification, one effect"
format:
  html:
    toc: true
    toc-depth: 2
    df-print: kable
execute:
  warning: false
  message: false
---

The previous chapter computed marginal effects for a single network. Sometimes the
question spans several — a few classes, schools or teams — and then what is wanted
is what the mechanism does across them, and how much they differ.

This chapter does that at the smallest size that still shows the whole arc: three
classes, one specification, one marginal effect. The next chapter does the same
for seventeen classes and six specifications, which adds bookkeeping rather than
new ideas, so this is the version to copy.

Defining effect sizes that can be compared across studies has a long history
outside network analysis, and the standard machinery is the random-effects or
multilevel model: a common part plus a study-specific part, with the option of
asking whether study-level variables — size, year, setting — account for the
differences.

For SAOMs this is already worked out. Section 12.3 of the RSiena manual
[@SienaManual] covers the practical side, and the two-stage population model is
discussed in more detail in @AMAND.

This chapter is in the comfortable case: the data are all to hand, so the same
specification can be fitted to every class, and each fit can be checked for
convergence before its estimate goes into the pool. That is worth noting because it
is not the usual case. Synthesising published results means taking the
specifications other authors chose, which will differ; and even here, fitting the
same specification everywhere does not make the classes identical in how
predictable behaviour is. Chapter 1 raised both points and Chapter 6 returns to
what a pooled estimate can be read as.

The pooling below uses `metafor` [@Viechtbauer2010], which is the general-purpose
tool for this and does not know anything about networks: it takes an estimate and
a standard error per study, whatever they came from.

```{r setup}
#| include: false
library(RSiena)
library(metafor)
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
CRE <- "#2166ac"   # creation
MNT <- "#d73027"   # maintenance
```

## The groups and their data

One entry per group. A "group" is whatever your studies are — here school
classes from the Dutch early-adolescence study of @Knecht2006, each with a
friendship network at two time points plus a gender covariate.

```{r data}
## The AK data are not shipped with this material -- see data/README.md.
## This downloads them into data/ the first time, and does nothing afterwards.
source("scripts/download_data.R")
ensure_ak_data()

load("data/AK_friendship.RData")   # friendship[[class]][["W"]], [["X"]]
load("data/AK_gender.RData")       # male[[class]]

groups <- c("12b", "12c", "12g")

make_group_data <- function(g) {
  n <- length(male[[g]])
  make_data_rsiena(
    friendship = as_dependent_rsiena(
      array(c(friendship[[g]][["W"]], friendship[[g]][["X"]]),
            dim = c(n, n, 2))),
    sex_m = as_covariate_rsiena(male[[g]], centered = FALSE)
  )
}

sapply(groups, function(g) length(male[[g]]))
```

## One specification for all of them

The same effects for every group, so that each contributes an estimate of the
same quantity. Density and reciprocity are included by default.

```{r spec}
make_group_spec <- function(dat) {
  eff <- make_specification(dat)
  eff <- set_effect(eff, transTrip1, verbose = FALSE)
  eff <- set_effect(eff, c(egoX, altX), covar1 = "sex_m", verbose = FALSE)
  eff
}
```

## The effect we want from each group

The first difference in the probability that a tie is present after one ministep,
when reciprocity goes from 0 to 1 — computed separately for tie creation and tie
maintenance, which is what `condition = "density"` does.

The margins are kept apart for the reason the previous chapter gave:
creation and maintenance are different decisions with quite different baseline
probabilities, so a single number averages over both.

```{r targets}
make_group_targets <- function(fit, eff) {
  tg <- make_marginal_targets(fit, effects = eff, depvar = "friendship",
                              type = "tieProb", level = "period",
                              condition = "density", includeDefaults = FALSE)
  set_target(tg, recip, contrast = c(0, 1), name = "recip_fd", verbose = FALSE)
}
```

## Estimate and compute, one group at a time

A plain loop, so you can stop after one group and look at it.

Worth keeping an eye on the convergence indicator: above roughly 0.25 the fit has
not converged, and it is better to rerun that group with `prevAns = fit` — as
`scripts/estimation.R` does — before its estimate goes into a pool.

```{r estimate}
algo      <- set_algorithm_saom(cond = FALSE, n3 = 1000, seed = 1234567)
per_group <- list()

for (g in groups) {
  dat <- make_group_data(g)
  eff <- make_group_spec(dat)
  fit <- siena(data = dat, effects = eff, control_algo = algo,
               batch = TRUE, verbose = FALSE)

  me <- marginalEffects(
    fit, dat, targets = make_group_targets(fit, eff),
    control_uncertainty = set_postest_uncertainty_saom(mode = "delta"))

  me <- as.data.frame(me)

  ## The conditioning adds one column named <depvar>_<statistic>; find it by
  ## suffix rather than by name, so this keeps working if the depvar is renamed.
  dens <- grep("_density$", names(me), value = TRUE)[1]

  me$group     <- g
  me$converged <- round(as.vector(fit$tconv.max), 3)
  me$margin    <- ifelse(me[[dens]] > 0, "creation", "maintenance")

  per_group[[g]] <- me[, c("group", "converged", "margin", "firstDiff", "SE")]
}

results <- do.call(rbind, per_group)
rownames(results) <- NULL
results$firstDiff <- round(results$firstDiff, 4)
results$SE        <- round(results$SE, 4)
results
```

Three classes, two margins each. Two things to notice before pooling: how far
apart the margins are within a single class, and how much the classes differ from
one another. Summarising both is what the next step is for.

## Pool

One random-effects meta-analysis per margin. The estimate and its standard error
are all `metafor` needs.

```{r pool}
pooled <- do.call(rbind, lapply(unique(results$margin), function(mg) {
  d <- subset(results, margin == mg)
  m <- rma(yi = d$firstDiff, sei = d$SE)
  data.frame(margin  = mg,
             est     = m$b[1, 1],
             ci_low  = m$ci.lb,
             ci_high = m$ci.ub,
             tau2    = m$tau2,   # between-group variance
             QEp     = m$QEp,    # small = groups differ more than chance
             k       = m$k)
}))
pooled[, -1] <- round(pooled[, -1], 4)
pooled
```

```{r pool-plot}
#| fig-width: 7
#| fig-height: 3
ggplot(results, aes(firstDiff, group, colour = margin)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = firstDiff - 1.96 * SE,
                      xmax = firstDiff + 1.96 * SE),
                  position = position_dodge(width = 0.4)) +
  geom_point(data = pooled, aes(est, "pooled", colour = margin),
             shape = 18, size = 4,
             position = position_dodge(width = 0.4)) +
  scale_colour_manual(values = c(creation = CRE, maintenance = MNT),
                      name = NULL) +
  labs(x = "effect of reciprocity on tie probability (pp)", y = NULL)
```

Across these classes, reciprocity shifts the probability that a tie is present
after one decision opportunity by the pooled estimate, in probability points. `tau2` says how much the classes genuinely differ; a small `QEp` says they
differ by more than sampling error alone would give.

Two things carried over from the previous chapter. This is a *per-opportunity*
quantity, describing one ministep rather than a whole period. And the interval is
analytic: it linearises the map from parameters to probabilities, so it comes out
symmetric around the estimate. A simulation-based interval propagates the curvature
of that map instead, which can show up where a probability sits close to zero or
one. Both rest on the same normal approximation for the parameter estimates that
`siena07` already makes for the coefficients, so what separates them is the
treatment of the map rather than of the parameters.

## Adapting this to your own study

Three things change, and nothing else has to:

1. **The groups and their data** — `groups`, and `make_group_data()` to build one
   RSiena data object per group from whatever you have.
2. **The specification** — `make_group_spec()`, the same effects for every group.
3. **The target** — `make_group_targets()`. Add further `set_target()` calls for
   more effects, or `set_second_diff()` for an interaction.

The loop, the convergence check and the pooling stay as they are.

## Where to go next

| you want | see |
|---|---|
| several specifications per group | `scripts/estimation.R` |
| the model-free comparison to set these against | `scripts/descriptives.R` |
| averaging over simulated ministeps, or accumulating over a period | `scripts/Postestimation.R` |
| pooling at scale, with plots per effect and specification | `scripts/meta_analysis.R`, and the next chapter |
