---
title: "Marginal effects across 17 school classes"
subtitle: "The same quantities as the worked example, applied and pooled"
format:
  html:
    toc: true
    toc-depth: 2
    df-print: kable
    code-fold: true
    code-summary: "show the code"
execute:
  warning: false
  message: false
---

## What was done

The quantities here are the ones the single-network walkthrough builds up:
predicted probabilities at a ministep, first and second differences, and the
split between tie creation and tie maintenance.

Here they are computed for **17 school classes**, drawn from the Dutch
early-adolescence study of @Knecht2006. Each class is fitted separately under six
specifications, the same post-estimation targets are computed for each, and the
per-class results are pooled with a random-effects meta-analysis — so every pooled
estimate below rests on `k = 17` classes, one observation per class.

The pipeline that produced all of the results, along with the
output:

| script | produces |
|---|---|
| `scripts/Postestimation.R` | per-class predicted probabilities, first and second differences |
| `scripts/meta_analysis.R` | the pooled tables and plots shown below |

Since the results are included too, you can pick up wherever suits: re-pool from
the saved per-class estimates in seconds, recompute those from the fitted models
in minutes, or go back to estimation and change the specifications.
`scripts/README.md` has the run order and what each stage costs.

The rest of this chapter walks through what came out.

```{r meta-load}
#| include: false
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
RES  <- "results"   # .rds produced by estimation / postestimation
OUTD <- "output"    # tables and plots produced by meta_analysis.R

files <- c(file.path(OUTD, c("table_raw_meta.csv", "table_static_meta.csv",
                             "table_accumulated_meta.csv",
                             "table_context_meta.csv")),
           file.path(RES,  c("postest_all_df.rds", "estimation_results.rds")))
have <- all(file.exists(files))
if (!have)
    message("missing: ", paste(files[!file.exists(files)], collapse = ", "))

get <- function(f) read.csv(file.path(OUTD, f))
```

```{r meta-missing}
#| echo: false
#| eval: !expr '!have'
cat("The pooled result files are not next to this document, so this section is
",
    "skipped.  They are produced by Postestimation.R and then
",
    "meta_analysis.R.
", sep = "")
```

## The three specifications

The full analysis fits six specifications per class. Three are shown here, chosen
so that each pair differs in one thing:

| | network effects | covariates |
|---|---|---|
| **model 0** | density, reciprocity, transitivity, reciprocity × transitivity | — |
| **model 2** | the same four | primary school, gender of ego and alter, and their interaction |
| **model 3** | density, reciprocity, transitivity | the same as model 2 |

So **model 0 → model 2** adds the covariates to a purely structural model, and
**model 2 → model 3** removes the reciprocity × transitivity interaction from the
full model. Reciprocity appears in all three, which is why it is the effect
followed below.

```{r keep-models}
#| include: false
KEEP <- c("model_0", "model_2", "model_3")
```

## Meta-analysis of average marginal effects

The walkthrough on single networks ends with **average marginal effects**: first
differences in the probability of a tie, averaged over decision situations. Those
are per-opportunity quantities — they describe one ministep.

Here is the AME for reciprocity across the 17 classes, next to the coefficient it
came from and to the accumulated effect introduced below:

- the **estimated coefficient**, as reported by `siena07`;
- the **average marginal effect (AME)**, the first difference in the probability
  of a tie averaged over decision situations — one ministep;
- the **accumulated effect**, the same contrast summed along the model-implied
  ministep sequence, so it refers to a whole period.

```{r meta-scales}
#| eval: !expr have
one <- function(f, eff, lbl) {
  d <- get(f); d <- d[d$effect == eff, c("model", "est")]
  names(d)[2] <- lbl; d
}
sc <- Reduce(function(x, y) merge(x, y, by = "model"),
             list(one("table_raw_meta.csv", "friendship_recip_eval",
                     "coefficient"),
                  one("table_static_meta.csv", "recip_fd",
                      "AME"),
                  one("table_accumulated_meta.csv", "recip_fd",
                      "accumulated")))
sc <- sc[sc$model %in% KEEP, ]
sc[, -1] <- round(sc[, -1], 3)
rownames(sc) <- NULL
sc
```

The three agree here on which specification gives the strongest reciprocity
effect. That is an observation about these data rather than something to count
on: the scaling arguments in Chapter 1 say a coefficient and its probability-scale
counterparts need not move together, and none of the three is adjusted for
whatever its own specification leaves out.

What they are differs. The coefficient is a difference in log-odds within a choice
set. The AME is a probability change per decision opportunity, averaged over
ministeps that are not themselves observed. The accumulated effect refers to a
whole period, but on a cumulative count scale rather than a probability one —
Chapter 6 sets out what that scale does and does not support.

What the probability scale does not fix on its own is the size of the choice set.
These classes run from 22 to 33 pupils, so the number of alternatives an actor
faces varies by about half across the pool, and the classes differ in composition
besides. Chapter 2's point applies here: these probabilities live under a $1/K$
geometry, so a per-opportunity probability of a given size does not mean quite the
same thing in a class of 22 as in one of 33. Moving off the coefficient scale buys
a common unit, not automatically a common baseline.

## Why the accumulated effect compares better across rates

The AME answers "how much per decision". It does not say how much happens over a
period, because it does not know how often actors decide — and they do not all
decide equally often. The rate parameter varies a lot between these classes:

```{r meta-rates}
#| eval: !expr have
raw <- readRDS(file.path(RES, "estimation_results.rds"))$coefs
p   <- readRDS(file.path(RES, "postest_all_df.rds"))

pick <- function(d, ...) { x <- subset(d, ...); x[, c("class_id", "est")] }
rt <- pick(raw, effect == "friendship_rate1" & model == "model_2")
so <- pick(p, estimand == "static" & condition == "pooled" &
              model == "model_2" & effect == "recip_fd")
pp <- pick(p, estimand == "accumulated" & condition == "pooled" &
              model == "model_2" & effect == "recip_fd")
names(rt)[2] <- "opportunities"; names(so)[2] <- "AME"
names(pp)[2] <- "accumulated"

cl <- merge(merge(rt, so), pp)
round(range(cl$opportunities), 2)
```

That matters, because the same effect per decision does not mean the same effect
over a period. Two classes make the point:

```{r meta-reorder}
#| eval: !expr have
cl <- cl[order(cl$AME), ]
cl[cl$class_id %in% c("12c", "02f"), ]
```

Class `12c` has roughly three times the reciprocity effect *per decision* of
class `02f`, yet **less** total movement over the period — because its actors
get about a third as many opportunities. On the per-opportunity scale `12c`
looks like the stronger case; on the period scale it is the weaker one.

Across all 17 classes the two scales rank the classes differently:

```{r meta-rank}
#| eval: !expr have
c(rank_correlation = round(cor(cl$AME, cl$accumulated,
                               method = "spearman"), 2),
  classes_that_change_rank = sum(rank(cl$AME) != rank(cl$accumulated)))
```

This is the specific sense in which the accumulated effect compares better: it
carries the rate, so it says what a mechanism does in a setting rather than what it
does per opportunity. Two networks running at different tempos can have identical
AMEs and quite different consequences over a period.

The price is interpretability. The accumulated effect is a sum, so it grows with
the number of opportunities and is not bounded the way a proportion is — a count of
extra ties created and dissolutions avoided rather than a probability. That makes
ratios between specifications readable and levels on their own much less so, and it
means the number should not be set against a between-wave difference in tie
proportions. Chapter 6 takes up those limits; the point here is only that when
opportunity rates differ, this is the scale on which the comparison can be made at
all.

## Does the choice of scale change the picture?

Mostly not, for the step from coefficient to probability. Each point below is one
class, placed by its **rank** on two scales; the diagonal is "no reordering at
all".

```{r scale-ranks}
#| eval: !expr have
#| fig-width: 9
#| fig-height: 5.5
raw <- readRDS(file.path(RES, "estimation_results.rds"))$coefs
p   <- readRDS(file.path(RES, "postest_all_df.rds"))

## Effect keys, paired with their raw-coefficient names and a readable label.
## The order here is the order the panels appear in, and the prose follows it:
## the two effects that behave alike first, the exception last.
eff_map <- data.frame(
  key   = c("recip_fd", "X_primary", "transTrip1_fd"),
  coefn = c("friendship_recip_eval", "friendship_X_primary_eval",
            "friendship_transTrip1_eval"),
  label = c("reciprocity", "primary school", "transitivity"))

ranks <- do.call(rbind, lapply(seq_len(nrow(eff_map)), function(i) {
  g <- function(d, ...) { x <- subset(d, ...); x[, c("class_id", "est")] }
  r <- g(raw, effect == eff_map$coefn[i] & model == "model_2")
  s <- g(p, estimand == "static" & condition == "pooled" &
            effect == eff_map$key[i] & model == "model_2")
  a <- g(p, estimand == "accumulated" & condition == "pooled" &
            effect == eff_map$key[i] & model == "model_2")
  names(r)[2] <- "coef"; names(s)[2] <- "ame"; names(a)[2] <- "acc"
  m <- merge(merge(r, s), a)
  rbind(
    data.frame(effect = eff_map$label[i], step = "coefficient → AME",
               x = rank(m$coef), y = rank(m$ame)),
    data.frame(effect = eff_map$label[i], step = "AME → accumulated",
               x = rank(m$ame),  y = rank(m$acc)))
}))

## Explicit level order for both facet dimensions.  Without this, facet_grid
## sorts alphabetically and puts "AME -> accumulated" above "coefficient ->
## AME", which is the reverse of the order the text reads them in.
ranks$step   <- factor(ranks$step,
                       levels = c("coefficient → AME",
                                  "AME → accumulated"))
ranks$effect <- factor(ranks$effect, levels = eff_map$label)

ggplot(ranks, aes(x, y)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = "dashed") +
  geom_point(colour = "#2166ac", size = 1.8, alpha = 0.85) +
  facet_grid(step ~ effect, switch = "y") +
  labs(x = "rank of the class on the earlier scale",
       y = "rank on the later scale") +
  theme(strip.placement  = "outside",
        strip.text.y.left = element_text(angle = 0, face = "bold"),
        strip.text.x      = element_text(face = "italic"),
        panel.spacing     = unit(0.9, "lines"))
```

Rows are the change of scale (bold, on the left), columns the effect (italic, on
top).

```{r scale-rho}
#| eval: !expr have
rho <- do.call(rbind, lapply(split(ranks, ~ effect + step, drop = TRUE),
  function(d) data.frame(step = as.character(d$step[1]),
                         effect = as.character(d$effect[1]),
                         rank_correlation = round(cor(d$x, d$y,
                                                      method = "spearman"), 2))))
rho <- rho[order(match(rho$step, levels(ranks$step)),
                 match(rho$effect, levels(ranks$effect))), ]
rownames(rho) <- NULL
rho
```

**Top row — coefficient to AME.** Rescaling a coefficient into a probability
leaves the classes in almost the same order for reciprocity and for the
primary-school effect: whatever the coefficient told you about which class has the
strongest effect, the probability tells you too. Transitivity is the exception.

**Bottom row — AME to accumulated.** This is the step just made, and here the
classes do reorder, because the question changes from what a mechanism does per
decision to what it does over a period.

### Which AME?

Both scales above pool tie creation and tie maintenance into one number. Chapters 2
to 4 kept them apart, so it is worth asking what the pooling costs here. The same
rank correlation, computed against the creation-only and maintenance-only averages:

```{r scale-ranks-margin}
#| eval: !expr have
margin_rho <- do.call(rbind, lapply(seq_len(nrow(eff_map)), function(i) {
  r <- subset(raw, effect == eff_map$coefn[i] &
                   model == "model_2")[, c("class_id", "est")]
  rho <- function(cond, st) {
    d <- subset(p, estimand == "static" & condition == cond &
                   effect == eff_map$key[i] & model == "model_2")
    if (!is.na(st)) d <- d[d$stratum == st, ]
    z <- merge(r, d[, c("class_id", "est")], by = "class_id")
    round(cor(z$est.x, z$est.y, method = "spearman"), 2)
  }
  data.frame(effect      = eff_map$label[i],
             pooled      = rho("pooled", NA),
             creation    = rho("bydensity",  1),
             maintenance = rho("bydensity", -1))
}))
rownames(margin_rho) <- NULL
margin_rho
```

Two things follow, and one does not.

The pooled column and the creation column are nearly identical. That is what a
sparse network implies — almost every alternative an actor faces is a tie that
could be created — so the pooled average is a creation average with a little
maintenance mixed in. Where this chapter says "the AME", read "mostly the creation
AME".

The maintenance column is where the coefficient stops predicting the ordering.
Reciprocity falls from 0.95 to 0.56 and the primary-school effect from 0.97 to
0.88. So the reassuring reading of the top row holds for the decision to *make* a
tie and not for the decision to *keep* one: which class has the strongest
maintenance effect is not something its coefficient reliably tells you. That is the
same lesson as the creation/maintenance split in the earlier chapters, arriving one
level up — about the ordering of classes rather than the size of an effect.

Transitivity is not explained away by either column. Its low correlation is not an
artefact of mixing margins: it is 0.43 on creation and 0.27 on maintenance, low
either way. Transitivity really is the effect whose class ordering changes when you
move off the coefficient scale.

## The result: a trade-off that only appears when keeping ties

With the scale settled, a substantive question. Do reciprocity and transitivity
reinforce each other, or substitute for one another? That is a **second
difference** — how much the effect of one more two-path changes when a tie goes
from unreciprocated to reciprocated — and everything below is on the accumulated
scale.

Start with it as it would normally be reported: one estimate per specification,
averaged over every ministep of the period.

```{r int-pooled}
#| eval: !expr have
#| fig-width: 7
#| fig-height: 2.2
pool <- get("table_accumulated_meta.csv")
pool <- pool[pool$effect == "interaction_sd" & pool$model %in% KEEP,
             c("model", "est", "ci_low", "ci_high")]

ggplot(pool, aes(est, model, xmin = ci_low, xmax = ci_high)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(colour = "grey25") +
  labs(x = "accumulated interaction over a period", y = NULL,
       title = "Pooled over all ministeps")
```

Small, negative, and easy to pass over. Reciprocity and transitivity look like
mild substitutes.

Now split the same estimate by the kind of decision the actor was facing, on the
same axis:

```{r int-split}
#| eval: !expr have
#| fig-width: 7
#| fig-height: 3
ctx <- get("table_context_meta.csv")
ctx <- ctx[ctx$effect == "interaction_sd" & ctx$estimand == "accumulated" &
           ctx$model %in% KEEP,
           c("model", "context", "est", "ci_low", "ci_high")]

both <- rbind(cbind(pool, context = "pooled"), ctx)
both$context <- factor(both$context,
                       levels = c("pooled", "creation", "maintenance"))
stopifnot(!anyNA(both$context))

ggplot(both, aes(est, model, xmin = ci_low, xmax = ci_high, colour = context)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  scale_colour_manual(values = c(pooled = "grey25",
                                 creation = "#2166ac",
                                 maintenance = "#d73027")) +
  labs(x = "accumulated interaction over a period", y = NULL, colour = NULL,
       title = "Split by the kind of decision")
```

```{r int-table}
#| eval: !expr have
ctx[, 3:5] <- round(ctx[, 3:5], 3)
ctx <- ctx[order(ctx$model, ctx$context), ]
rownames(ctx) <- NULL
ctx                     # model 3 has no interaction to report
```

The pooled point is not a compromise between the two — it sits almost on top of
the creation value, because creation decisions vastly outnumber maintenance ones
and dominate the average.

The pooled estimates hide how much the classes differ. All 17 of them, with the
meta-analytic summary at the bottom:

```{r int-classes}
#| eval: !expr have
#| fig-width: 9
#| fig-height: 6.5
per <- readRDS(file.path(RES, "postest_all_df.rds"))
per <- per[per$estimand == "accumulated" & per$condition == "bydensity" &
           per$effect == "interaction_sd" & per$model %in% KEEP &
           per$stratum != 0, ]
per$context  <- ifelse(per$stratum > 0, "creation", "maintenance")
per$ci_low   <- per$est - 1.96 * per$se
per$ci_high  <- per$est + 1.96 * per$se

summ <- ctx
summ$class_id <- "Summary"

fig <- rbind(per[, c("class_id", "model", "context", "est", "ci_low", "ci_high")],
             summ[, c("class_id", "model", "context", "est", "ci_low", "ci_high")])
fig$class_id <- factor(fig$class_id,
                       levels = c("Summary", sort(unique(per$class_id),
                                                  decreasing = TRUE)))
fig$context  <- factor(fig$context, levels = c("creation", "maintenance"))

ggplot(fig, aes(est, class_id, xmin = ci_low, xmax = ci_high, colour = context,
                shape = class_id == "Summary")) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(position = position_dodge(width = 0.6), fatten = 1.6) +
  facet_wrap(~ model) +
  scale_colour_manual(values = c(creation = "#2166ac",
                                 maintenance = "#d73027")) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 18), guide = "none") +
  ## a few single-class intervals run past +/- 3; clipped so the rest is legible
  coord_cartesian(xlim = c(-2.5, 1)) +
  labs(x = "accumulated second difference over a period", y = NULL,
       colour = NULL)
```

Individual classes are noisy: most intervals cross zero, which is about what one
class of thirty pupils supports, and a few run far past the edges of the plot.
No single class would settle the question.

What survives pooling is the ordering. The maintenance estimate is below the
creation estimate in **33 of the 34** class-by-specification combinations, and
the two summary intervals do not overlap. The result is not driven by a handful
of classes; it is a small, consistent difference repeated across all of them,
which is the case for pooling in the first place.

Reciprocity and transitivity substitute for one another when an actor decides
whether to **keep** a tie, and do essentially nothing when deciding whether to
**make** one. These are not a strong and a weak version of one finding: one is
absent, the other is large. Reported as a single number, the maintenance result
disappears entirely.

Equivalent figures for every specification and every effect ship in `output/`,
written by `scripts/meta_analysis.R`. To do this on your own data, start from
Chapter 4, which is the same arc at a size that can be copied.

## In short

- The coefficient, the AME and the accumulated effect agree here on direction and
  respond similarly to the specification — an observation about these data rather
  than something to rely on in general.
- Only the accumulated effect carries the rate, which is what lets it be compared
  between settings whose actors decide at different frequencies. It is a
  cumulative count rather than a probability, so it is read through ratios rather
  than as a between-wave difference.
- Splitting by kind of decision changes the substantive conclusion rather than
  refining it: the reciprocity × transitivity second difference is absent in tie
  creation and large in tie maintenance.
